# Assessment for Working with Data

## *Nikita Sridhar \\* 10/20/23

*Loading necessary data and libraries*

```{r}
library(tidyverse)
library(here)

bor_nestlings <- read_csv(here::here("data/73_species.csv"))

bor_traits <- read_csv(here::here("data/Traits_73_species.csv"))

avonet <- readxl::read_excel(here::here("data/AVONET Supplementary dataset 1.xlsx"), "AVONET1_BirdLife")
```

*Are nestlings showing up earlier in the year over time? - all code provided by Max as an example*

```{r}
all_birds_trend <- bor_nestlings %>% 
  group_by(Year) %>% 
  summarize(mean_doy = mean(Dayofyear))

ggplot(all_birds_trend, aes(Year, mean_doy)) + 
  geom_point() +
  geom_smooth(method = "lm")

#breakdown by species
species_trends <- bor_nestlings %>% 
  group_by(Year, Species) %>% 
  summarize(mean_doy = mean(Dayofyear),
            .groups = "drop")

ggplot(species_trends, aes(Year, mean_doy, color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm")

#focusing on 5 most data-rich species
data_richness <- bor_nestlings %>% 
  count(Species)

most_rich <- data_richness %>% 
  arrange(desc(n)) %>% 
  slice(1:5)

most_rich_trends <- bor_nestlings %>% 
  filter(Species %in% most_rich$Species) %>% 
  group_by(Species, Year) %>% 
  summarize(mean_doy = mean(Dayofyear), 
            .groups = "drop")

ggplot(most_rich_trends, aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")

#Find slope of lines to determine which species have strongest trends - following function is to find slope between x and y
trend <- function(x, y) {
  xy_lm <- lm(y ~ x)
  coef(xy_lm)[2]
}

# Calculate the trend for all species
bor_trends <- species_trends %>% 
  group_by(Species) %>% 
  summarize(doy_trend = trend(Year, mean_doy))

#spot check two species
soi <- c("ARDCIN", "LARMIN")
species_trends %>% 
  filter(Species %in% soi) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

*My turn - visualizing species with the most positive and most negative trends*

```{r}

nrow_bor_trends <- nrow(bor_trends) # Use this later
bor_extreme <- bor_trends %>% 
  # Sort by the day of year trend
  arrange(desc(doy_trend)) %>% 
# Keep just the first (most negative trend) and last (most positive trend) rows
  slice(c(1,73))

# Now plot them
species_trends %>% 
  filter(Species %in% bor_extreme$Species) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) + 
    geom_point() +
    geom_smooth(method = "lm")
```

*Seeing whether smaller birds have more flexible phenology (absolute value of trend is correlated with size)*

```{r}
# First, select and rename important columns
avonet_size <- select(avonet, SciName = Species1, Mass_g = Mass)
bor_sciname <- select(bor_traits, 
                      Species = Abbreviation, 
                      SciName = `Scientific name`)

#joining things
bor_trends_size <- bor_trends %>% 
  left_join(bor_sciname, by = "Species") %>% 
  left_join(avonet_size, by = "SciName") %>% 
  mutate(abs_trend = abs(doy_trend))

# Plot it
ggplot(bor_trends_size, aes(Mass_g, abs_trend)) +
  geom_point() +
  geom_smooth(method = "lm")

#removing missing values
bor_trends_size2 <- bor_trends_size %>%
  drop_na(Mass_g) %>%
  filter(Mass_g < 2000, abs_trend < 1)

ggplot(bor_trends_size2, aes(Mass_g, abs_trend)) +
  geom_point() +
  geom_smooth(method = "lm")
```

The above graph shows that our hypothesis is not supported - the magnitude of the trend for the day of the year birds show up and bird body size (mass) are not correlated.
